Revisiting divisible, injective and flabby abelian groups from a constructive point of view

### 2022-09-27 slides-fischbachau2022.tex
Slides finished for tomorrow

\usepackage{graphbox}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd,wasysym}
\usepackage{bussproofs}
\usepackage{xspace}
\usepackage{amssymb}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{ifthen}
\usepackage{bbm}
\usepackage[normalem]{ulem}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning}
\hypersetup{colorlinks=true}

@@

\begin{frame}{The double-negation embedding}
  \vspace*{-0.8em}
  \mbox{How to extract \good{constructive proofs} from
  \bad{classical proofs} as the following?}

@@

  \textbf{Thm.} Let~$X$ be an inhabited set of natural numbers.
  \begin{enumerate}
    \item If~\badbox{\textsc{lem}}, then~$X$ contains a minimal element.
    \item<5-> If~$X$ is \hil{detachable}, then~$X$ contains a minimal element.
    \item<7-> It is \notnot the case that~$X$ contains a minimal element.
  \end{enumerate}

@@

@@

>>>>

\usepackage{graphbox}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd,wasysym}
\usepackage{bussproofs}
\usepackage{xspace}
\usepackage{amssymb}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{ifthen}
\usepackage[normalem]{ulem}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning}
\hypersetup{colorlinks=true}

@@

\begin{frame}{A case study in double negation}
  \vspace*{-0.8em}
  \mbox{How to extract \good{constructive proofs} from
  \bad{classical proofs} as the following?}

@@

  \textbf{Thm.} Let~$X$ be an inhabited set of natural numbers.
  \begin{enumerate}
    \item Assuming~\badbox{\textsc{lem}}, the set~$X$ contains a minimal element.
    \item<5-> If~$X$ is \hil{detachable}, then~$X$ contains a minimal element.
    \item<7-> It is \hil{not~not} the case that~$X$ contains a minimal element.
  \end{enumerate}

@@

\newcommand{\hnegneg}{\hil{$\neg\neg$}}
\newcommand{\negg}{{\neg}\!\!\!{\neg}}
\newcommand{\neggv}{{\neg}\!\!\!\phantom{{\neg}}}
\newcommand{\bott}{\bot\!\!\!\!\bot}
\newcommand{\bottv}{\bot\!\!\!\!\phantom{\bot}}

\begin{frame}{The double-negation embedding}
  \justifying
  \vspace*{-1em}
  \small
  \textbf{Def.} For formulas over a fixed first-order signature, the
  \hil{$\neg\neg$-translation} $\varphi \mapsto \varphi^{\neg\neg}$ is defined
  by the following clauses.
  \begin{align*}
    (\varphi_\text{atomic})^{\neg\neg} &\defeqv \hnegneg \varphi_\text{atomic} &
    (\varphi \Rightarrow \psi)^{\neg\neg} &\defeqv (\varphi^{\neg\neg} \Rightarrow \psi^{\neg\neg}) \\
    \bot^{\neg\neg} &\defeqv \hnegneg\bot &
    \top^{\neg\neg} &\defeqv \top \\
    (\varphi \vee \psi)^{\neg\neg} &\defeqv \hnegneg(\varphi^{\neg\neg} \vee \psi^{\neg\neg}) &
    (\varphi \wedge \psi)^{\neg\neg} &\defeqv (\varphi^{\neg\neg} \wedge \psi^{\neg\neg}) \\
    (\exists x\?X\_ \varphi)^{\neg\neg} &\defeqv \hnegneg{(\exists x\?X\_ \varphi^{\neg\neg})} &
    (\forall x\?X\_ \varphi)^{\neg\neg} &\defeqv (\forall x\?X\_ \varphi^{\neg\neg}) \\
  \end{align*}
  \vspace*{-3em}

  \mbox{\textbf{Ex.} $(\forall a{:}X\_\! \exists b{:}X\_\! a = b \vee \ldots)^{\neg\neg}
  \equiv (\forall a{:}X\_\! \hnegneg\exists b{:}X\_\! \hnegneg(\hnegneg(a=b) \vee
  (\ldots)^{\neg\neg})).$}
  \pause

  \textbf{Prop.} Classically, $\varphi \Leftrightarrow \varphi^{\neg\neg}$.
  \pause
  \vspace*{-0.5em}

  \textbf{Thm.} For every formula~$\varphi$ and set of formulas~$\Gamma$:
  \vspace*{-0.6em}
  \begin{enumerate}
    \item Minimally, $\neg\neg(\varphi^{\neg\neg}) \Rightarrow
    \varphi^{\neg\neg}$.
    \item Minimally, $\varphi^{\neg\neg} \Leftrightarrow \neg\neg\varphi$ in
    case that~$\varphi$ is geometric
    ($R{\top}{\bot}{\wedge}{\vee}{\exists}{\bigvee}$).
    \item Classical logic proves~$\varphi$ from~$\Gamma$
    iff minimal logic proves~$\varphi^{\neg\neg}$ from~$\Gamma^{\neg\neg}$.
  \end{enumerate}
  \pause
  \vspace*{-0.3em}

  \textbf{Cor.} If PA proves~$\varphi$, then~HA proves~$\varphi^{\neg\neg}$.
  \pause

  \textbf{Rem.} Theorem and corollary hold for every \hil{local
  operator}~$\nabla$ in place of~$\neg\neg$, in particular for~$\negg\negg\varphi \defeq
  ((\varphi \Rightarrow \bott) \Rightarrow \bott)$ for some arbitrary
  formula~$\bott$.
\end{frame}

\newcommand{\negx}{\hil{${\only<1>{\neggv}\only<2->{\negg}}$}}
\newcommand{\botx}{\hil{${\only<1>{\bottv}\only<2>{\bott}\only<3->{\beta}}$}}

\begin{frame}{Barr's theorem / Friedman's trick / A-translation}
  \textbf{Thm.} Let~$\Gamma$ be a set of geometric sequents over a fixed
  signature. Let~$\sigma$ be a geometric sequent. Then the following are
  equivalent:
  \begin{enumerate}
    \addtocounter{enumi}{-1}
    \item $\sigma$ holds for the \good{generic model} of~$\Gamma$ (in its
    classifying topos).
    \item $\sigma$ is provable from~$\Gamma$ in \good{geometric logic}.
    \item $\sigma$ is provable from~$\Gamma$ in \good{intuitionistic logic}.
    \item $\sigma$ is provable from~$\Gamma$ in \bad{classical logic}.
    \item (Assuming \badbox{\textsc{ac}}) $\sigma$ is provable from~$\Gamma$ in
    \bad{classical logic with~\textsc{ac}}.
  \end{enumerate}

  \emph{Proof of~``\,\,$\normalnumber{3} \Rightarrow \normalnumber{2}$''.}
  Write~$\sigma \equiv (\alpha \seq{\vec x} \beta)$. Then intuitionistically,
  \[ \alpha \Longrightarrow
    \negx\negx\alpha \Longleftrightarrow
    \alpha^{\negx\negx} \Longrightarrow
    \beta^{\negx\negx} \Longleftrightarrow
    \negx\negx\beta \equiv
    ((\beta{\Longrightarrow}\botx){\Rightarrow}\botx)
    \visible<4->{\Longrightarrow \beta.}
  \]
\end{frame}

\section{Constructive proofs from invalid proofs}

@@

\begin{tabular}{ll@{\qquad\qquad}ll}
  \toprule
  $\varphi$ & $\varphi^{\neg\neg}$ &
  $\varphi$ & $\varphi^{\neg\neg}$ \\\midrule
  $\alpha_\text{atomic}$ & $\hnegneg \alpha_\text{atomic}$ &
  $\alpha \Rightarrow \beta$ & $\alpha^{\nabla} \Rightarrow \beta^{\nabla}$ \\
  $\bot^{\neg\neg}$ & $\hnegneg\bot$ &
  $\top^{\neg\neg}$ & $\top$ \\
  $\alpha \vee \beta$ & $\hnegneg(\alpha^{\neg\neg} \vee \beta^{\neg\neg})$ &
  $\alpha \wedge \beta$ & $\alpha^{\neg\neg} \wedge \beta^{\neg\neg}$ \\
  \bottomrule
\end{tabular}
